\documentclass[Minh_PhD_thesis.tex]{subfiles}
\begin{document}
% Use unnumbered section for acknowledgements
\section*{ACKNOWLEDGEMENTS}
\phantomsection

% Add reference to the table of contents {toc} at the section level {section} titled "Acknowledgements" {Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

I would like to express my deepest gratitude to my supervisor, Dr. Sorin Dr\u{a}ghici, for his invaluable guidance, continuous support, and encouragement throughout my research journey. His insightful feedback and unwavering patience has been instrumental in shaping this dissertation and my growth as a researcher.

I am also profoundly grateful to my parents for their unconditional love, support, and sacrifices that have made this journey possible. Their belief in me has been a constant source of motivation.

To my wife, Truc Bui, her patience, understanding, and unwavering support have been my greatest strength. Your encouragement has kept me going through the challenges of this journey, and for that, I am forever thankful.

I would also like to extend my appreciation to my current and former lab mates, especially Tin Nguyen, Adib Shafi, Douglas Craig, Cristina Mitrea and Radu Vanciu, whose collaboration, insightful discussions, and camaraderie have made this experience both enriching and enjoyable. Their support and shared knowledge have greatly contributed to this work.

Finally, I acknowledge everyone who has contributed directly or indirectly to the completion of this dissertation. Thank you for all of your support and encouragement.

%Thank you.

\end{document}
